---
title: "Assignment 3: Bayesian Analysis  \nLighthouse Problem\n"
author: "Shane Weisz (WSZSHA001)"
date: "19/09/2019"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
header-includes: \usepackage{float} \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
---
This assignment involves the use of Bayesian Analysis methods to solve 'The Lighthouse Problem', where we attempt to draw inference on the coordinates of a lighthouse at an unknown location in the sea given a series of flashes emitted by the lighthouse that are observed at different points on the coastline.

## (i) Derivation of posterior distribution of $\alpha,\beta|x$

We want to derive the posterior distribution of $\alpha,\beta|\boldsymbol{x}$, which can be expressed as:
$$\pi(\alpha,\beta|\boldsymbol{x}) = \frac{\pi(\alpha,\beta) \text{L}(\alpha,\beta;\boldsymbol{x})}{m(\boldsymbol{x})}= \frac{g(\alpha,\beta;\boldsymbol{x})}{m(\boldsymbol{x})},$$ where 
$g(\alpha,\beta;\boldsymbol{x})=\pi(\alpha,\beta) \text{L}(\alpha,\beta;\boldsymbol{x})$.

#### Prior: $\pi(\alpha,\beta)$

From the context of the problem, we can make the assumption that $\alpha$ and $\beta$ are independent. As such, we can obtain our prior distribution from $$\pi(\alpha,\beta)=\pi(\alpha)\pi(\beta)$$. 

We are given that $\alpha \sim \text{U}(0,5)$ and $\beta \sim \text{Gamma}(4,8)$, and so substituting the appropriate density functions yields our **prior distribution**: $$\pi(\alpha,\beta)=\pi(\alpha)\pi(\beta) = \frac{1}{5}\frac{8^4 \beta^3 e^{-8\beta}}{\Gamma(4)}=\frac{2048\beta^3 e^{-8\beta}}{15}$$

#### Likelihood: $\text{L}(\alpha,\beta$;$\boldsymbol{x}$)

We first note that we calculate the likelihood as follows:
$$\text{L}(\alpha,\beta, \boldsymbol{x})=\prod_{i=1}^{200} f(x_i|\alpha,\beta)$$

We thus need to calculate the density function for the data, $f(x_i|\alpha,\beta)$. 
So let $\theta_i$ be the angle of inclination of the line joining the lighthouse to the point on the coastline where the $i^{th}$ emission was recorded. It follows that $\tan(\theta_i)=\frac{x_i-\alpha}{\beta}$ and hence that $\theta_l=-\arctan(\frac{\alpha}{\beta})<\theta_k<\arctan(\frac{5-\alpha}{\beta})=\theta_u$.

Now the problem definition allows us to make the assumption that $\theta_i \sim\text{U}(\theta_l,\theta_u)$, which yields the p.d.f. of $\theta_i$ as $\pi(\theta_i)=\frac{1}{\theta_u-\theta_l}=\frac{1}{\text{A}(\alpha,\beta)}$, where we define $\text{A}(\alpha,\beta) = \theta_u-\theta_l$.

Applying an appropriate univariate transformation, we obtain that $f(x_i|\alpha,\beta)=\frac{1}{\text{A}(\alpha,\beta)}\begin{vmatrix}\frac{d \theta_i}{d {x_i}}\end{vmatrix}$ where $\frac{d \theta_i}{d {x_i}}=\frac{d}{d {x_i}}\arctan(\frac{x_i-\alpha}{\beta})=\frac{\beta}{\beta^2+(x_k-\alpha)^2}$
Since $\beta>0$, we get that $\frac{\beta}{\beta^2+(x_k-\alpha)^2}>0$, and so  $f(x_i|\alpha,\beta)=\frac{1}{\text{A}(\alpha,\beta)}\cdot\frac{\beta}{\beta^2+(x_i-\alpha)^2}$.

Substituting into the above formula for the likelihood gives us our **likelihood** function:
$$\text{L}(\alpha,\beta, \boldsymbol{x})=\prod_{i=1}^{200} f(x_i|\alpha,\beta) = \prod_{i=1}^{200} \frac{1}{\text{A}(\alpha,\beta)}\cdot\frac{\beta}{\beta^2+(x_i-\alpha)^2}$$

#### Predictive pdf of x: $m(\boldsymbol{x})$

By the Law of Total Probability, we have that $$m(\boldsymbol{x})=\int_{0}^5 \int_{0}^{\infty} \pi(\alpha,\beta) \text{L}(\alpha,\beta;\boldsymbol{x}) d\beta d\alpha$$ where $\pi(\alpha,\beta)=\frac{2048\beta^3 e^{-8\beta}}{15}$ and $\text{L}(\alpha,\beta, \boldsymbol{x})= \prod_{i=1}^{200} \frac{1}{\text{A}(\alpha,\beta)}\cdot\frac{\beta}{\beta^2+(x_i-\alpha)^2}$

## (ii) Function to calculate g($\alpha$,$\beta$;x) and obtain contour plot


```{r loadData, include=FALSE}
load("~/Documents/UCT/3rd Year/STA3043S/Assignments/Assignment3/Assignment3Data2019.RData")
x_data = Data['dckmat004',]
```

The following functions below are used to calculate g($\alpha$,$\beta$;x), by means of helper functions A and L.

```{r gfunc}
# A(alpha, beta) is defined as the difference between the upper and lower bounds for theta
A = function(alpha, beta)
{
   return( (atan((5 - alpha) / beta)) - (-atan(alpha / beta)) )
}

# Returns the likelihood function for alpha and beta given an observed data vector x
L = function(alpha, beta, x)
{
   return( prod( (1 / A(alpha, beta)) * beta / (beta ^ 2 + (x - alpha)^2)) )
}

# Returns g(alpha,beta;x) - the numerator of the posterior
gfunc = function(alpha, beta, x)
{
    return( dunif(alpha, 0, 5) * dgamma(beta, shape = 4, rate = 8) * L(alpha, beta, x) )
}
gfunc.vec = Vectorize(gfunc, vectorize.args = c("alpha","beta"))
```

We use the above function to obtain the following Contour Plot of the joint posterior distribution of the
two parameters with the $\alpha$-axis ranging between (0, 5) and the $\beta$-axis ranging between (0, 1).

```{r contourPlot, echo=FALSE, fig.cap="Contour Plot"}
n = 100
alpha_range = seq(from = 0, to = 5, length.out = n)
beta_range  = seq(from = 0, to = 1, length.out = n)
g = outer(alpha_range, beta_range, FUN = gfunc.vec, x_data)

contour(x = alpha_range, 
        y = beta_range, 
        z = g, 
        xlab=expression(alpha),
        ylab=expression(beta),
        main=expression(paste("Proportional to ", pi, "(", alpha, ",", beta, "|x)")), 
        xlim=range(alpha_range), 
        ylim=range(beta_range), 
        col="blue",
        nlevels = 10)
```

## (iii) Values of $\alpha$, $\beta$ that maximises $\pi$($\alpha$,$\beta$|x)

The appropriate values are **2.559** for alpha and **0.842** for beta.

```{r maximise, echo=TRUE}
# A function that returns a value proportional to the negative log of the posterior
negLogPost = function(params, x)
{
    alpha = params[1]
    beta  = params[2]
    return( -1 * log( (dunif(alpha, 0, 5) * dgamma(beta, 4, 8) * L(alpha, beta, x)) ) )
}

estimates = optim( par = c(2.5, 0.5), 
                   fn = negLogPost, 
                   x = x_data,
                   lower = c(0.0001, 0.0001),
                   upper = c(5, Inf),
                   method = "L-BFGS-B")$par
# names(estimates) = c('alpha', 'beta')

library(knitr)
library(kableExtra)
headings = c('','$\\alpha $', '$\\beta$')
estimates_mtx <- matrix(c('Estimate', round(estimates,3)), ncol = length(estimates) + 1)
colnames(estimates_mtx) <- c(headings)
kable(estimates_mtx, format = 'latex', align="c", escape = F, caption = "Estimates for $\\alpha$ and $\\beta$")  %>% kable_styling(latex_options = c("hold_position"))  
```

## (iv) Function to obtain marginal distribution of $\alpha|x$ using numerical integration

```{r alpham}

```

## (v) Estimates of the posterior mean and variance of $\alpha$

```{r estimates}

```

## (vi) Acceptance-rejection method for inference on $\alpha$

```{r ar}

```

